Capturing regeneration in models of forest dynamics

### 2022-11-17 manuscript/manuscriptRW.Rmd
add mortality figures and change label "empirical" with "observed" in figures script and manuscript

Species simulated: 4C only simulated 4 species. Treemig. ForClim and
Empirical data have more species assigned all together into category
"others". LPJ-GUESS simulated bushes that were not consider in the
output of speces composition neither recruitment.

@@

ToDo add a map with the species composition of the empirical data

@@

Results related to regeneration levels, species composition and
mortality in recruitment varied across models and in comparison with the
observations. As expected, recruitment was overestimated across most
models for both the 7 and 10 cm diameter thresholds, with the exception
of SYBILA, Landis II and aDGVM2 that estimated recruitment rates at the
lower end of the plausibility interval of the empirical data (Figure
\@ref(fig:over)). The models with highest overestimation were PICUS and
TreeMig. For most of the models, the variability across sites in
recruitment levels (visible from the interquartile range in the box
plots of Figure \@ref(fig:over)) is similar to or smaller than the
observed variability for most of the models, with the exception of
PICUS, where the mean recruitment variability is much larger.

@@

When looking at the representation of recruitment rates along key
gradients of light availability (basal area), temperature (degree-days)
and soil moisture (water balance), distinct features emerge: the models
reproduce the effect of basal area much better than the effects of
climatic gradients, where they feature varying effects (Figure
\@ref(fig:nvTrend710BA), \@ref(fig:nvTrend710BA500) and FigureS 1-4).

@@

Most of the models match the level of diversity (as measured by the
Shannon index) of the observed data quite well (Figure \@ref(fig:H710)),
although they consistently underestimate its variability across sites.
Four models overestimate recruitment diversity (4C, PICUS, LANDIS-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case as this model simulated 5 species only. In most
models there are small differences across sites regarding the species
richness of recruitment between 7 and 10 cm. Most models maintain or
even increase in recruitment diversity between the 7 and 10 cm
thresholds, and the same is visible from the empirical data, however the
differences are not significant. **EXPLAIN WHY  "Can we understand this
(I know this is for the Discussion, but still...)?? This cannot be
species richness (as there are no new species popping between 7 and 10
cm DBH). I'd also expect the number of species to be lower at 10 cm than
at 7 cm (because some will die off due to insufficient light etc.), but
this is not reflected in H' because species with zero abundance are
ignored (ln[0] is not defined). So a strongly increasing evenness must
be driving this increase of H'."**

@@

```{r diff, echo=FALSE, fig.cap="Departure of each model in the representation of the main species recruitment relative abundance from the observations across the environmental gradients for a 7cm treshold. The values shown are the mean of samples per site and across sites in bins, where the range of the climatic water balance (cm) and seasonal degree-day sum (dºC) was divided in 10 bins. The point sizes represent ratio between recruited basal area of the species and the total recruited basal area for all the species. The color gradient shows the difference between the simulated and observed ratio between recruited basal area of the species and the total recruited basal area: grey cell represent areas of the envirnomental gradient in which the model had R BA share  and the empirical dara did not becuase both recruitment and stand level recruitments were equal to zero.", out.width='100%'}
knitr::include_graphics(here::here("figures", "HeatEnvGrad_10circles_7.png"))
```

@@

**Main outcomes mortality in number of trees:**

-   There are differences across models on the mortality rate between 7
    and 10 cm (Figure \@ref(fig:mortality11)).

-   Most of the models have a mortality rate larger than those present
    in the empirical data (Figure \@ref(fig:mortality11)).

-   Several models simulate a mortality rate in recruitment that could
    be explained by the self thinning rule (Figure
    \@ref(fig:mortality11)).

-   Models below the Reineke line have lower mortality than the expected
    by self thinning rule, which might point to facilitation or higher
    mortality before the species reached 7 cm diameter (Figure
    \@ref(fig:mortality11)).

-   Values below 1 line show sites where there is a higher number of 10
    cm trees than in 7 cm- this is an artifact of the method - we expect
    space for time substitution, but in reality we just have one
    measurement (Figure \@ref(fig:mortality11)).

-   Models such as SYBILA, xComp, PICUS, LandClim amd TreeMig are above
    the self thinning line (Reineke) indicating that they have higher
    mortality in recruitment of mixed species forest than the expected
    in an even aged single species forests (Figure
    \@ref(fig:mortality11)).

-   Most of the models with lower variation in mortality between 7 and
    10cm also have lower overestimation in recruitment at 7cm (Figure
    \@ref(fig:mortality13) and FigureS 13).

-   Models with higher mortality also show higher tree recruitment at
    7cm (Figure \@ref(fig:mortality13)).

-   There might be a difference between models including stochastic
    processes in recruitment and those which don't, but perhaps the
    reality implies higher mortality rates in the adult component
    (enhance the share of less shade tolerant species?) [to the
    discussion] ((Figure \@ref(fig:mortality11) and
    \@ref(fig:mortality13)).

-   Some of the models that start at a lower number of individuals (7cm)
    have a lower decrease between 7 and 10cm dbh (ratio 7-10cm). The
    species competition and self-thinning is either not that large or it
    happened before the species reached 7cm. This pattern it is not
    consistent across models (FigureS 14-15)

-   There is an effect of the number of individuals at 7cm on the
    mortality rate. Most of the models show a positive effect
    (increasing recruitment at 7cm shows increases the mortality between
    7 and 10cm), but some models show a negative effect. The cloud point
    per models shows the differences across sites, the shape point cloud
    per model shows if the models take into account regeneration density
    when applying mortality to regeneration (FigureS 14-15) This is not
    that clear anymore when we calculate the slope of the linear trends
    per model.

```{r  mortality11, echo=FALSE, fig.cap="Ratio of tree recruitment between 7 and 10 cm recruitment treshold. Dashed blue lines mark a ratio equal to 1 indicating no decrease between 7 and 10cm tree recruitment and ratio equal to 1.77 corresponding with the Reineke selfthinning ratio under evenaged conditions", out.width='75%'}
knitr::include_graphics(here::here("figures", "ratio7_10.png"))
```

```{r mortality13, echo=FALSE, fig.cap="Ratio of tree recruitment between 7 and 10 cm recruitment treshold and the overestimation proportion (simulated - empirical / empirical) for the mean recruitment trees per model across sites and samples. The horizontal dotted lines show a ratio equal to 1 indicating no decrease between 7 and 10cm tree recruitment and ratio equal to 1.77 corresponding with the Reineke selfthinning ratio under evenaged conditions. The vertical line representes the 0 overestimation proportion", out.width='100%'}
knitr::include_graphics(here::here("figures", "mort7_10_medianOver.png"))
```

@@

>>>>

Species simulated: 4C only simulated 4 species. Treemig. ForClim and
Observed data have more species assigned all together into category
"others". LPJ-GUESS simulated bushes that were not consider in the
output of speces composition neither recruitment.

@@

ToDo add a map with the species composition of the observed data

@@

Results related to regeneration levels, species composition and
mortality in recruitment varied across models and in comparison with the
observations. As expected, recruitment was overestimated across most
models for both the 7 and 10 cm diameter thresholds, with the exception
of SYBILA, Landis II and aDGVM2 that estimated recruitment rates at the
lower end of the plausibility interval of the observed data (Figure
\@ref(fig:over)). The models with highest overestimation were PICUS and
TreeMig. For most of the models, the variability across sites in
recruitment levels (visible from the interquartile range in the box
plots of Figure \@ref(fig:over)) is similar to or smaller than the
observed variability for most of the models, with the exception of
PICUS, where the mean recruitment variability is much larger.

@@

When looking at the representation of recruitment rates along key
gradients of light availability (basal area), temperature (degree-days)
and soil moisture (water balance), distinct features emerge: the models
reproduce the effect of basal area much better than the effects of
climatic gradients, where they feature varying effects (Figure
\@ref(fig:nvTrend710BA) and FigureS 1-4).

@@

Most of the models match the level of diversity (as measured by the
Shannon index) of the observed data quite well (Figure \@ref(fig:H710)),
although they consistently underestimate its variability across sites.
Four models overestimate recruitment diversity:4C, PICUS, LANDIS-II and
LPJ-GUESS, the latter mostly for the 10 cm diameter threshold. The model
4C is a special case as this model simulated 5 species only. In most
models there are small differences across sites regarding the species
richness of recruitment between 7 and 10 cm. Most models maintain the
recruitment diversity between the 7 and 10 cm thresholds, and the same
is visible from the observed data, there are only few models where the
recruitment diversity increases between 7 and 10 cm which might be an
artifact of the sample strategy and calculating the Shannon index based
in BA as for a same sample recruitment at 10cm will include larger trees
with larger BA and if those are from different species, the Shannon
index will be larger.

@@

```{r diff, echo=FALSE, fig.cap="Departure of each model in the representation of the main species recruitment relative abundance from the observations across the environmental gradients for a 7cm treshold. The values shown are the mean of samples per site and across sites in bins, where the range of the climatic water balance (cm) and seasonal degree-day sum (dºC) was divided in 10 bins. The point sizes represent ratio between recruited basal area of the species and the total recruited basal area for all the species. The color gradient shows the difference between the simulated and observed ratio between recruited basal area of the species and the total recruited basal area: grey cell represent areas of the envirnomental gradient in which the model had R BA share  and the observed data did not becuase both recruitment and stand level recruitments were equal to zero.", out.width='100%'}
knitr::include_graphics(here::here("figures", "HeatEnvGrad_10circles_7.png"))
```

@@

There are strong differences across models on the mortality rate between
7 and 10 cm (Figure \@ref(fig:mortality11)). All models except for the
two DGVMs LPJ-GUESS and aDGVM2 feature a mortality rate larger than the
observed data, thus at least partially compensating for the general
overestimation of recruitment rates (Figure \@ref(fig:over)). Some of
the models showed strong compensation for the initial overestimation of
recruitment with larger mortality of recruitment between 7 and 10 cm
thresholds but not in others. On the one hand, models below the Reineke
line have lower mortality than expected by the self-thinning rule, which
may point to facilitation, or simply a higher mortality rate before the
trees reached 7 cm diameter. On the other hand, models such as xComp,
PICUS and TreeMig are above the self-thinning line (Reineke), indicating
that they have higher mortality in recruitment of mixed species forest
than the expected in an even-aged single species forests, which
indicates strong compensation for the initial overestimation of
recruitment in some models. However, some of the models simulate a
mortality rate in recruitment that is broadly compatible with the
self-thinning rule, i.e. 4C, SIBYLA , LandClim and LANDIS II.

Models with higher mortality also show higher tree recruitment at 7cm
(Figure \@ref(fig:)\@ref(fig:mortality13)). On the other hand, most of
the models with a low variation in mortality between 7 and 10 cm
(ForCEEPS, ForCLim 1, ForClim 11 and LandClim) also have lower
overestimation in recruitment at 7 cm (Figure \@ref(fig:mortality11) and
FigureS XX). Some of the models that start at a lower number of
individuals (7cm thershold) have a lower decrease between 7 and 10cm dbh
(ratio 7-10cm); in these models the species competition and
self-thinning is either not that large or it happened before the species
reached 7cm. However, this pattern it is not consistent across models
(FigureS XX). There is an effect of the number of individuals at 7cm on
the mortality rate. Most of the models show a positive effect
(increasing recruitment at 7cm shows increases the mortality between 7
and 10cm), but some models show a negative effect, however the trend it
is not very high and not always significant when we look at the slope of
the linear trends per model (FigureS XX and tableS XX).

```{r  mortality11, echo=FALSE, fig.cap="Ratio of tree recruitment rates between the 7 and 10 cm thresholds. Dashed blue lines mark a ratio equal to 1, indicating no decrease in tree recruitment between 7 and 10 cm, and a ratio equal to 1.77, corresponding to the Reineke self-thinning ratio under even-aged conditions.", out.width='75%'}
knitr::include_graphics(here::here("figures", "ratio7_10.png"))
```

```{r mortality13, echo=FALSE, fig.cap="Ratio of tree recruitment between 7 and 10 cm recruitment treshold and the overestimation proportion (simulated - observed / observed) for the mean recruitment trees per model across sites and samples. The horizontal dotted lines show a ratio equal to 1 indicating no decrease between 7 and 10cm tree recruitment and ratio equal to 1.77 corresponding with the Reineke selfthinning ratio under evenaged conditions. The vertical line represents the 0 overestimation proportion", out.width='100%'}
knitr::include_graphics(here::here("figures", "mort7_10_medianOver.png"))
```

@@

-   \# H1: the higher regen7 the stronger the regen mortality

    \# decrease in stem number between regen7 and regen10

    \# as ratio nn7/nn10 (colname nn710)

    \# reference lines

    \# -ratio 1 = no decrease from 7 -\> 10

    \# -ratio for selfthinning under even aged conditions

    \# with Reineke exp rk=-1.605: (7/10)\^(-1.605) = 1.77

-   Models below the Reineke line have lower mortality than expected by
    the self-thinning
    rule, which may point to facilitation, or simply a higher mortality
    rate before the trees reached 7 cm diameter. But then these models
    would need to underestimate recruitment levels (Fig. 1), no? But
    this is not what we see... not sure how to explain.

-   methodological limitations compare mortality with the self thinning
    line etc Values below the 1.0 line show sites where there is a
    higher number of 10 cm trees than in 7 cm - this is an
    artifact of the method - we expect space for time substitution, but
    in reality
    we just have one measurement (Figure 7).

-   There might be a difference between models including stochastic
    processes in recruitment and those which do not, but perhaps the
    reality implies higher mortality rates in the adult component
    (enhance the share of less shade tolerant species?) [to the
    discussion] ((Figure \@ref(fig:mortality11) and
    \@ref(fig:mortality13)).

-    Most models maintain or even increase in recruitment diversity
    between the 7 and 10 cm thresholds, and the same is visible from the
    observed data, however the differences are not significant.
    **EXPLAIN WHY  "Can we understand this (I know this is for the
    Discussion, but still...)?? This cannot be species richness (as
    there are no new species popping between 7 and 10 cm DBH). I'd also
    expect the number of species to be lower at 10 cm than at 7 cm
    (because some will die off due to insufficient light etc.), but this
    is not reflected in H' because species with zero abundance are
    ignored (ln[0] is not defined). So a strongly increasing evenness
    must be driving this increase of H'."** This could be because we are
    comparing 7 and 10cm from same samples, and at 10cm you might
    include more trees and at 7cm This calls for further research and
    anotehr paper :D
